---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
terra::terraOptions(progress = 0)
```

# bsrmap: Biosecurity Risk Mapping

<!-- badges: start -->
[![Last commit](https://img.shields.io/github/last-commit/cebra-analytics/bsrmap.svg)](https://github.com/cebra-analytics/bsrmap/commits/main)
<!-- badges: end -->

The *bsrmap* package provides a collection of functions for estimating the spatial distribution of risk, or establishment likelihood, of exotic pests, diseases, and other biosecurity threats, and is based on *edmaps* (Camac & Baumgartner, 2021) and its associated research and consultative development, which is described in Camac et al. (2020; 2021) and Camac (2024). The risk mapping functions provide methods for transforming and combining data or spatial layers to estimate pest or threat suitability, pathway arrival likelihoods, and the overall establishment likelihood. The function collection includes:

1. *abiotic_suitability*: builds a suitability layer for abiotic factors, such as environment & climate, often via a Species Distribution Model (SDM - see <https://github.com/cebra-analytics/bssdm>)
1. *biotic_suitability*: builds a suitability layer for biotic factors, such as threat host, often by combining vegetation, land-use, and other landscape features suitable for the threat
1. *pest_suitability*: combines abiotic and biotic suitability layers to estimate an overall pest or threat suitability
1. *pathway_likelihood*: builds a likelihood layer based on (relative) arrival likelihood layers and threat border leakage and establishment viability estimates
1. *arrival_likelihood*: combines pathway likelihood layers to estimate an overall threat arrival likelihood
1. *establishment_likelihood*: combines threat (pest) suitability and arrival likelihood layers to estimate an overall threat establishment likelihood (or risk)
1. *aggregate_categories*: builds a layer or mask from selected classes from a categorical layer, such as vegetation or land-use, that are indicative of threat suitability or arrival likelihood
1. *buffered_hull_layer*: calculates a spatial layer based on fitting a alpha or convex hull with a distance buffer to a series of locations that are indicative of threat suitability or arrival likelihood
1. *combine_layers*: combines layers that are indicative of threat suitability or arrival likelihood
1. *conform_layer*: conforms the spatial configuration (CRS, extent, and resolution) of a spatial layer to that of a common configuration for the risk map via a template layer
1. *cumulative_likelihood*: calculates the ordered cumulative likelihood of a spatial likelihood layer
1. *distance_weight_layer*: calculates a spatial layer by applying a distance decay using proximity to a series of weighted locations that are indicative of threat suitability or arrival likelihood
1. *distribute_features*: calculates a spatial layer by distributing threat suitability or arrival values from shape polygons (e.g. land management regions) across a raster layer
1. *points_to_layer*: converts points to raster layers by assigning each layer value to the nearest point within a maximum range
1. *transform_layer*: transforms a spatial layer by applying a linear, exponential, logarithmic, or power function, or substituting values using lower or upper thresholds

Note that functions 1-6 provide the functionality for building the core components of a risk map. Functions 7-15 provide useful utilities for building the layers required for the core components.

## Installation

You can install the latest version of *bsrmap* from [GitHub](https://github.com/) with:

``` r
remotes::install_github("cebra-analytics/bsrmap")
```

## Example

The following example generates the establishment likelihood for a Hawkweed species (*Hieracium pilosella*), an exotic weed for Australia. We build the core components of our risk map at each step.

### Step 1: Abiotic suitability

The abiotic suitability for the threat generally refers to environmental factors, such as climate. Here we will utilise the *Range bagging* (Drake, 2015) SDM predicted climate suitability from our *bssdm* package example (see <https://github.com/cebra-analytics/bssdm>), which utilised global climate data from *WorldClim* (Fick & Hijmans, 2017; <http://www.worldclim.org>) and Hawkweed occurrence records from Global Biodiversity Information Facility (GBIF, 2026).

```{r example_1_1, message = FALSE, fig.align = "center"}
# Range bagging SDM predicted global climate suitability
sdm_predicted <- terra::rast("data/sdm_predicted_suitability.tif")
terra::plot(sdm_predicted, colNA = "grey",
            main = "Range bagging SDM predicted climate suitability",
            xlab = "Longitude (degrees)", ylab = "Latitude (degrees)")
```

The area of interest for our risk map is Australia only, which we will define using an Australian Albers (equal-area CRS) 1km-grid (resolution) template. Note that we will continue to use this template when building other components of our risk map so that layers may be consistently combined.

```{r example_1_2, message = FALSE, fig.align = "center"}
# Area of interest: Australian Albers CRS 1km-grid template
area_template <- terra::rast("data/template_au_1km.tif")
terra::plot(area_template, colNA = "grey",
            main = "Australian Albers 1km-grid template")
```

We will now conform our global SDM output to our area of interest template via the *abiotic_suitability* function.

```{r example_1_3, message = FALSE, fig.align = "center"}
# Abiotic suitability (conform to area of interest template)
abiotic_suitability <- bsrmap::abiotic_suitability(sdm_predicted,
                                                   area_template)
terra::plot(abiotic_suitability, main = "Abiotic suitability", colNA = "grey")
```

### Step 2: Biotic suitability

The biotic suitability of the threat generally refers to habitat or host requirements. Here we focus on vegetation and land-use types that are suitable for the Hawkweed. We firstly select suitable major vegetation groups (MVG) from National Vegetation Information System (NVIS) V7.0 (NVIS, 2025) raster layers, and build a binary suitability mask using our *aggregate_categories* function. Users may download [NVIS Raster Geodatabases](https://www.dcceew.gov.au/environment/environment-information-australia/national-vegetation-information-system/data-products) and place them in a suitable directory (e.g. *downloaded_data*) before loading and transforming the NVIS layer.

```{r example_2_1, message = FALSE, fig.align = "center"}
# Suitable NVIS major vegetation groups (MGV)
nvis_rast <- terra::rast(
  paste0("../downloaded_data/NVIS_V7_0_AUST_RASTERS_EXT_ALL/",
         "NVIS_V7_0_AUST_EXT.gdb"))[["NVIS7_0_AUST_EXT_MVG_ALB"]]
head(terra::cats(nvis_rast)[[1]][,1:2])
all_cats <- terra::cats(nvis_rast)[[1]][,1]
# Exclude 'closed' and aquatic vegetation categories
selected_cats <- all_cats[which(!all_cats %in% c(1,2,15,23,24,28))]
# Aggregate and binarise selected categories
biotic_nvis_suitability <- bsrmap::aggregate_categories(
  nvis_rast, area_template,
  categories = all_cats, selected = selected_cats,
  binarize = TRUE)
terra::plot(biotic_nvis_suitability, colNA = "grey",
            main = "Suitable NVIS major vegetation")
```

Next we select suitable (or exclude unsuitable) land-use types from Catchment Scale Land Use of Australia (ABARES, 2024) raster layers, which may be downloaded from [Australian Bureau of Agricultural and Resource Economics and Sciences (ABARES)](https://doi.org/10.25814/2w2p-ph98) and placed in a suitable directory. Again, we will use our *aggregate_categories* function to build another binary suitability mask for the Hawkweed.

```{r example_2_2, message = FALSE, fig.align = "center"}
# Suitable land use categories
land_use_rast <-
  terra::rast("../downloaded_data/clum_50m_2023_v2/clum_50m_2023_v2.tif")
head(terra::cats(land_use_rast)[[1]][,c(1, 3)])
all_cats <- terra::cats(land_use_rast)[[1]][,1]
# Exclude aquatic land-use categories
selected_cats <- all_cats[which(all_cats < 600)]
# Aggregate and binarise selected categories
biotic_land_use_suitability <- bsrmap::aggregate_categories(
  land_use_rast, area_template,
  categories = all_cats,
  selected = selected_cats,
  binarize = TRUE)
terra::plot(biotic_land_use_suitability, colNA = "grey",
            main = "Suitable ABARES land use")
```

Finally we can combine our vegetation and land use suitability mask layers into an overall biotic suitability layer via our *biotic_suitability* function. Note that we combine our layers via the "union" function, since a location (grid-cell) is suitable if either the vegetation or land-use is suitable.

```{r example_2_3, message = FALSE, fig.align = "center"}
# Combine (union) biotic suitability layers
biotic_suitability <- bsrmap::biotic_suitability(
  terra::rast(list(biotic_nvis_suitability, biotic_land_use_suitability)),
  use_fun = "union")
terra::plot(biotic_suitability, main = "Biotic suitability", colNA = "grey")
```

### Step 3: Threat suitability

We can now calculate the overall threat suitability by combining the abiotic and biotic suitability layers via our *pest_suitability* function. Note that this time we combine our layers via the "prod" function, since a location (grid-cell) is only suitable if both the abiotic and biotic conditions are suitable.

```{r example_3, message = FALSE, fig.align = "center"}
# Combine (multiply) abiotic and biotic suitability layers
threat_suitability <- bsrmap::pest_suitability(
  abiotic_suitability, biotic_suitability,
  use_fun = "prod")
terra::plot(threat_suitability, main = "Threat suitability", colNA = "grey")
```

### Step 4: Pathway likelihoods

The likelihood of the threat arriving in area of interest via one or more pathways may be calculated via our *pathway_likelihood* function, which builds each likelihood layer based on (relative) arrival likelihood layers and threat border leakage and establishment viability estimates. Here will calculate the threat arrival likelihoods for three different pathways:

1. Arrivals via agricultural products
1. Arrivals via tourists
1. Arrivals via mail

#### Agricultural products pathway

Arrivals via agricultural products can be estimated using data pertaining to the distribution of fertiliser use across Australian Natural Resource Management (NRM) regions (Australian Bureau of Statistics, 2016-17). This data was collated and transformed into shape-file layers by CEBRA (as described in Camac et al., 2020). The fertiliser data values for each NRM can distributed via our *distribute_features* function across an agricultural land-use raster layer, again calculated via our *aggregate_categories* using Catchment Scale Land Use of Australia (ABARES, 2024) raster layers. The resultant spatially-weighted layer can then used in our *pathway_likelihood* function to spatially distribute region-wide arrival likelihood estimates, which are generated within the function using estimated threat border leakage and establishment viability parameters (95% confidence intervals) for the pathway.

```{r example_4_1, message = FALSE, fig.align = "center"}
# Aggregate and binarise agricultural land use categories
agriculture_land_use <- bsrmap::aggregate_categories(
  land_use_rast, area_template,
  categories = all_cats,
  selected = c(330, 331, 332, 333, 334, 335, 336, 337, 338,
               340, 341, 342, 343, 344, 345, 346, 347, 348, 349, 350,
               351, 352, 353, 420, 421, 423, 424, 430, 431, 432, 433,
               434, 435, 436, 437, 438, 439, 440, 441, 442, 443, 444,
               445, 446, 447, 448, 449, 450, 451, 452, 453, 454, 510,
               511, 512, 513, 514),
  binarize = TRUE)
terra::plot(agriculture_land_use, colNA = "grey",
            main = "ABARES agricultural land use")
# Fertiliser use NRM distribution
fertiliser_nrm <- terra::vect("data/fertiliser_nrm.gpkg")
# Distribute fertiliser use across agricultural land use
fertiliser_distr <- bsrmap::distribute_features(
  agriculture_land_use, fertiliser_nrm,
  vars = "Fert_t")
terra::plot(fertiliser_distr, colNA = "grey",
            main = "Fertiliser use distributed across agricultural land use")
# Agricultural products pathway likelihood
agriculture_likelihood <- bsrmap::pathway_likelihood(
  pathway_layers = fertiliser_distr,
  leakage_rate_ci = c(3, 30),
  viability_rate_ci = c(0.001, 0.05),
  confidence = 0.95)
terra::plot(log(agriculture_likelihood + 1e-9, base = 10), colNA = "grey",
            main = "Agricultural products pathway likelihood (log)")
```

#### Tourists pathway

Arrivals via tourists can be estimated using data pertaining to the distribution of tourist accommodation (Australian Bureau of Statistics, 2015-16). The distribution data was collated and transformed into a raster layer by CEBRA (as described in Camac et al., 2020). We firstly conform this layer to our area of interest template via our *conform_layer* function. We then weight the distribution of tourist accommodation by their distance from major airports via our *distance_weight_layer* function, such that 50% of the likelihood is distributed within 500 km. The resultant spatially-weighted layer can then used in our *pathway_likelihood* function to spatially distribute region-wide arrival likelihood estimates, which are generated within the function using estimated threat border leakage and establishment viability parameters (95% confidence intervals) for the pathway.

```{r example_4_2, message = FALSE, fig.align = "center"}
# Conform the tourist accommodation distribution layer to template
tourist_distr <- bsrmap::conform_layer(
  terra::rast("data/tourist_accommodation.tif"), area_template)
terra::plot(log10(tourist_distr + 1), colNA = "grey",
            main = "Tourist accommodation (log)")
# Accommodation distribution weighted by distance from airport
airports_df <- read.csv("data/airports.csv")
distance_weight <- bsrmap::distance_weight_layer(
  area_template, airports_df,
  beta = log(0.5)/500,
  weights = NULL)
terra::plot(distance_weight, colNA = "grey",
            main = "Airport distance weight (500km half-decay)")
# Tourists pathway likelihood
tourists_likelihood <- bsrmap::pathway_likelihood(
  pathway_layers = terra::rast(list(tourist_distr, distance_weight)),
  leakage_rate_ci = c(1, 10),
  viability_rate_ci = c(0.001, 0.05),
  confidence = 0.95)
terra::plot(log(tourists_likelihood + 1e-7, base = 10), colNA = "grey",
            main = "Tourists pathway likelihood (log)")
```

#### Mail pathway

Arrivals via mail can be estimated using data pertaining to the distribution of human population in Australia (Australian Bureau of Statistics, 2024-25). We firstly conform this layer to our area of interest template via our *conform_layer* function. The resultant spatially-weighted layer can then used in our *pathway_likelihood* function to spatially distribute region-wide arrival likelihood estimates, which are generated within the function using estimated threat border leakage and establishment viability parameters (95% confidence intervals) for the pathway.

```{r example_4_3, message = FALSE, fig.align = "center"}
# Conform the human population distribution layer to template
population_distr <- bsrmap::conform_layer(
  terra::rast("data/australian_population_2025.tif"), area_template)
terra::plot(log10(population_distr + 1), colNA = "grey",
            main = "Human population (log)")
# Mail pathway likelihood
mail_likelihood <- bsrmap::pathway_likelihood(
  pathway_layers = population_distr,
  leakage_rate_ci = c(1, 10),
  viability_rate_ci = c(0.3, 0.9),
  confidence = 0.95)
terra::plot(log(mail_likelihood + 1e-8, base = 10), colNA = "grey",
            main = "Mail pathway likelihood (log)")
```

### Step 5: Arrival likelihood

We can now calculate the overall arrival likelihood by combining our three pathway likelihood layers via our *arrival_likelihood* function. Note that this time we combine our likelihood layers via the "union" function, which is calculated via $1-\prod_{i=1}^n(1-p_i)$ at each location (grid-cell) using the local arrival likelihoods $p_i$ for each $i$ of $n$ pathways. For small likelihood values the "union" function is approximately equivalent to adding the likelihood values. For example:

```{r example_5_1, message = FALSE, fig.align = "center"}
cell = 12975568
p_1 = agriculture_likelihood[cell][,1]
p_2 = tourists_likelihood[cell][,1]
p_3 = mail_likelihood[cell][,1]
1 - (1 - p_1)*(1 - p_2)*(1 - p_3)
p_1 + p_2 + p_3
```

Let's now calculate the overall arrival likelihood layer.

```{r example_5_2, message = FALSE, fig.align = "center"}
# Combine pathway arrival likelihood layers
arrival_likelihood <- bsrmap::arrival_likelihood(
  terra::rast(list(agriculture_likelihood,
                   tourists_likelihood,
                   mail_likelihood)),
  use_fun = "union")
arrival_likelihood[cell][,1] # compare with "union" example
terra::plot(log(arrival_likelihood + 1e-9, base = 10), colNA = "grey",
            main = "Arrival likelihood (log)")
```

### Step 6: Establishment likelihood

Finally, we can now calculate the overall threat establishment likelihood by combining the threat suitability and arrival likelihood layers via our *establishment_likelihood* function. Note that this time we combine our layers via the "prod" function, since for the threat to establish at a location (grid-cell), the location must be sufficiently suitable, and likely to arrive at the location (assuming reasonable independence between suitability and arrival probabilities).

```{r example_6, message = FALSE, fig.align = "center"}
# Combine threat suitability and arrival likelihood layers
establishment_likelihood <- bsrmap::establishment_likelihood(
  threat_suitability, arrival_likelihood,
  use_fun = "prod")
terra::plot(log(establishment_likelihood + 1e-12, base = 10), colNA = "grey",
            main = "Establishment likelihood (log)")
```

## References

ABARES (2024). 'Catchment Scale Land Use of Australia – Update December 2023 version 2', *Australian Bureau of Agricultural and Resource Economics and Sciences*, Canberra, June, CC BY 4.0, <doi:10.25814/2w2p-ph98>

Australian Bureau of Statistics (Reference period: 2015-16 financial year) 'Tourist accommodation - Australia' [data set], Tourist Accommodation, Australia. *ABS*. <https://www.abs.gov.au/statistics/industry/tourism-and-transport/tourist-accommodation-australia/latest-release>

Australian Bureau of Statistics (Reference period: 2016-17 financial year) 'Fertiliser use, Australia, year ended 30 June 2017' [data set], Land Management and Farming in Australia. *ABS*. <https://www.abs.gov.au/statistics/industry/agriculture/land-management-and-farming-australia/latest-release>

Australian Bureau of Statistics (Reference period: 2024-25 financial year) 'Australian population grid 2025 in GeoTIFF format' [data set], Regional population. *ABS*. <https://www.abs.gov.au/statistics/people/population/regional-population/latest-release>

Camac, J. S., Baumgartner, J. B., Robinson, A., & Elith, J. (2020). 'Developing pragmatic maps of establishment likelihood for plant pests'. Tech. Rep. 170607, *Centre of Excellence for Biosecurity Risk Analysis (CEBRA), The University of Melbourne, Melbourne*.
<https://cebra.unimelb.edu.au/__data/assets/pdf_file/0012/3539397/170607_final_report.pdf>

Camac, J. & Baumgartner, J. (2021). '*edmaps* (early detection maps) : An R package for creating Australian maps of establishment likelihood for terrestrial plant pests'. R package <https://github.com/jscamac/edmaps>

Camac, J. S., Baumgartner, J. B., Hester, S., Subasinghe, R., & Collins, S. (2021). 'Using edmaps & Zonation to inform multi-pest early detection surveillance designs'. Tech. Rep. 20121001, *Centre of Excellence for Biosecurity Risk Analysis (CEBRA), The University of Melbourne, Melbourne*. <https://cebra.unimelb.edu.au/__data/assets/pdf_file/0009/3889773/20121001_final_report.pdf>

Camac, J. S. (2024). 'Detect: Designing post-border surveillance schemes. In Hester et al. (Eds.), Biosecurity: A Systems Perspective. *Taylor & Francis*. <doi:10.1201/9781003253204>

Drake, J. M. (2015). 'Range bagging: a new method for ecological niche modelling from presence-only data'. *Journal of the Royal Society Interface*, 12(107), 20150086. <doi:10.1098/rsif.2015.0086>

Fick, S. E., Hijmans, R. J. (2017). 'WorldClim 2: new 1-km spatial resolution climate surfaces for global land areas'. *International Journal of Climatology*, 37, 4302–4315. <doi:10.1002/joc.5086>

GBIF.org (04 May 2026) 'GBIF Occurrence Download'. <doi:10.15468/dl.q6q6fk>

National Vegetation Information System (NVIS) V7.0 <https://www.dcceew.gov.au/> from copyright Commonwealth of Australia 2025
